---
title: "Report of twin transition in SMEs across Europe: analysis from Eurobarometer data"
author: "Giulia Picardi"
date: "2025-06-08"
output: html_document
bibliography: elementi_esportati.json 
csl: apa.csl

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

prova citazione (@horbachDigitalisationSustainabilityMeasures2024)

```{r}

```


# Introduction

The transition toward a greener and more digital economy has become a defining challenge for the European Union. In response to climate change, resource scarcity, technological disruption, and long-term strategic goals, the EU has placed the twin transition - the simultaneous advancement of digitalisation and environmental sustainability - at the core of its policy agenda (European Commission, 2020). This dual transformation requires the active involvement of all sectors of society, including businesses, governments, and civil society. Among economic actors, small and medium-sized enterprises (SMEs) are considered a central pillar of the European productive system and are expected to play a crucial role in translating policy objectives into concrete operational practices. Yet, the capacity of SMEs to pursue both digital and environmental strategies remains highly uneven, depending on structural conditions, organisational capabilities, and available incentives.
Although some firms have started to integrate digital and green strategies, the degree of integration varies significantly across the SME landscape. The literature has examined this relationship from multiple perspectives, including eco-innovation and corporate social responsibility (CSR) (Horbach, 2024; Triguero et al., 2013; del Río et al., 2016). Digital technologies are increasingly viewed not only as tools for improving efficiency, but also as catalysts for systemic change. These technologies enable real-time monitoring and tracking, support life-cycle assessment and predictive simulations, and facilitate the virtualisation of activities - thereby reducing emissions linked to production, logistics, and mobility (JRC, 2022). However, the relationship between digitalisation and sustainability is not deterministic. Organisational factors such as firm size, internal competences, and strategic orientation play a key role in shaping whether and how these two dimensions are pursued jointly, especially within SMEs (Horbach, 2024).
This study investigates the association between digital activity and environmental engagement among European SMEs using data from the Flash Eurobarometer 486 (2020), which provides cross-sectional insights into firm behaviour across EU member states. Since the survey predates both the adoption of the European Green Deal and the onset of the COVID-19 pandemic, it offers a unique snapshot of SME strategies at a time when sustainability was gaining traction, but before the implementation of major policy and financial instruments. This temporal context allows for the analysis of structural conditions that shaped firms' engagement in the twin transition prior to institutional acceleration.

Drawing on the model developed by Horbach (2024), the analysis assesses whether digitally active firms are more likely to implement environmental measures, treating digitalisation as a potential enabler of green practices. This approach is informed by a broader body of research on the uneven diffusion of sustainability innovations (Triguero et al., 2013; del Río et al., 2016), and the temporal lags that often exist between policy design and behavioural change at the firm level (OECD, 2019). 
The purpose of this analysis is to contribute to a deeper understanding of how the twin transition is implemented in practice - and to identify the organisational conditions that may support its implementation within the European SME landscape. 
To guide the selection of variables, this study follows the literature-based framework adopted by Horbach (2024), who identifies relevant indicators of digitalisation and environmental engagement by drawing on previous research on eco-innovation and digital transformation in SMEs. Digitalisation is captured through the presence of specific technological applications—such as AI, cloud computing, big data analytics, and automation—while environmental engagement is assessed through firm-level practices including energy efficiency, resource savings, recycling, and the development of sustainable products.



# Methods

## Data

The data used are thriven from the Flash Eurobarometer 486, a survey conducted among small and medium-sized enterprises (SMEs) in all EU Member States in February 2020. The sample includes firms across all industrial sectors, stratified by country, size, and industry. We restrict our analysis to the 27 EU countries and focus on firms with complete information on key variables.

Dependent Variables
Five binary indicators representing green sustainability practices are selected following Horbach's methods:

•	eco_innovation (Q19_5)
•	recycling (Q24_1)
•	resource_reduction (Q24_2)
•	energy_saving (Q24_3)
•	sustainable_products (Q24_4)

Each is coded as 1 if the company reports engaging in the respective activity, 0 otherwise. Additionally, a sustainability_index (range 0–5) is constructed by summing these five indicators, providing a count measure of environmental engagement.


Independent Variables

The main explanatory variables are six binary indicators of digital technology adoption:
•	AI (Q23_1)
•	Big data (Q23_5)
•	Cloud computing (Q23_2)
•	High-speed infrastructure (Q23_6)
•	Robotics (Q23_3)
•	Smart devices/sensors (Q23_4)

Each variable equals 1 if the firm reports using the technology.


Control Variables

Following Horbach, firm-level controls capturing economic, structural, and contextual characteristics, are included:

•	Export orientation (Q12B_1–3)
•	Family ownership (Q13_7)
•	Financial capacity (Q4A_7–8)
•	High-growth firm (Q5_1_4, Q5_2_4)
•	Industry location (Q8_4)
•	Founding year (before 2000) (Q1)
•	Skill shortage as sustainability barrier (Q26_4)
•	Firm size (number of employees) (Q2B)
•	Urban location (Q8_1)


## Statistical Models

Two complementary models:

1.	Multivariate Probit: jointly models the probability of engaging in each green action, accounting for correlations between the binary outcomes

2.	Negative Binomial: models the sustainability index (0–5) using the same covariates, to account for overdispersion in the count data



# Results
tbd
```{r}

```

# Discussion
tbd

## Bibliography 










