---
title: "Twin transition in SMEs across Europe: a firm-level analysis from Eurobarometer data"
author: "Giulia Picardi"
date: "05 july 2025"
output: 
  bookdown::html_document2:
  toc: true
  number_sections: true
  link_citations: true
bibliography: multivariate.json 
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

```
# Abstract

The European twin transition agenda places SMEs at the centre of efforts to combine digitalisation and environmental sustainability. This paper tests the hypothesis that European SMEs adopting digital technologies are more likely to implement green-oriented practices. Cross-sectional data from Flash Eurobarometer 486 (2020) are used to estimate a multivariate probit model on five ecological activities: eco-innovation, recycling, resource efficiency, energy saving, and the development of sustainable products.
The analysis explicitly focuses on environmental outcomes of sustainability – excluding broader CSR dimensions - to refine the scope of inquiry in relation to prior studies and isolate the ecological core of the twin transition. This focus leverages the strength of the multivariate specification, which estimates correlated behaviours within a unified domain. The results suggest that cloud computing, big data, and robotic automation are consistently associated with green engagement, particularly in innovation and efficiency-related areas. Conversely, AI and smart devices display weaker and more selective association. Skill shortages appear negatively associated with energy-saving and resource-reduction behaviours, pointing to human capital as a potential constraint in the ecological transformation of SMEs. 

While the analysis does not allow for causal claims, it highlights patterned alignments between digital capabilities and ecological practices. By focusing on a pre-policy juncture, it offers a grounded perspective on how SMEs engaged with environmental objectives in the absence of structured institutional mandates - shedding light on the early organisational dynamics of green alignment and providing a baseline for future research on how such trajectories evolve under shifting institutional conditions.


# Introduction

The transition toward a greener and more digital economy has become a defining challenge for the European Union. In response to climate change, resource scarcity, technological disruption, and long-term strategic goals, the EU has placed the twin transition as foundational principles of the future European economy [@EuropeanCommission2020]. This dual transformation requires the active involvement of all sectors of society, including businesses, governments, and civil society. 
Among economic actors, small and medium-sized enterprises are considered a central pillar of the European productive system and are expected to play a crucial role in translating policy objectives into concrete operational practices. These expectations can be interpreted as institutional pressures that shape not only the strategic priorities of firms, but also the symbolic alignment with broader societal goals. Yet, the capacity of SMEs to pursue both digital and environmental strategies remains highly uneven, depending on structural conditions, organisational capabilities, and available incentives. Although some firms have started to integrate digital and green strategies, the degree of integration varies significantly across the SME landscape. Despite the strategic prioritisation of the twin transition at EU level, many SMEs face structural and cognitive barriers that hinder the effective implementation of both green and digital practices [@RupeikaApoga2022]. 
The literature has examined this relationship from multiple perspectives, including eco-innovation and corporate social responsibility (CSR) [@Horbach2024; @TrigueroMorenoDavia2013; @DelrioPenascoRomero2016]. Digital technologies are increasingly viewed not only as tools for improving efficiency, but also as catalysts for systemic change. These technologies enable real-time monitoring and tracking, support life-cycle assessment and predictive simulations, and facilitate the virtualisation of activities - thereby reducing emissions linked to production, logistics, and mobility [@JRC2023].

Following DiMaggio and Powell’s [@DiMaggio1983] theory of institutional isomorphism, the diffusion of digital and green practices among firms may not solely reflect instrumental adaptation to market or environmental challenges, but also a process of conformity to institutionalised expectations. As the EU consolidates a normative agenda linking innovation and sustainability, organisations may adopt digital and environmental routines to gain legitimacy within their fields, regardless of short-term efficiency gains. From this perspective, the alignment between digitalisation and sustainability is not merely a matter of capabilities or incentives, but also a response to the symbolic pressures embedded in the European policy discourse. 
This lens is particularly relevant to our study, which tests whether SMEs reporting digital adoption are also more likely to engage in ecological practices. If isomorphic pressures are at play, one would expect a pattern of relational alignment between digitalisation and environmental engagement - even in the absence of formal “twin transition” programmes.
Rather than indicating a generalised twin transition across the SME landscape, the results reveal differentiated alignments between specific digital technologies and environmental outcomes, suggesting that institutional signals alone - without policy shifts and financial supports - were insufficient to generate systematic twin transition behaviours among SMEs.
Indeed, the relationship between digitalisation and sustainability is not deterministic.
Recent studies have pointed out that, although SMEs are central to the EU's twin transition strategy, the actual implementation of integrated digital and green strategies remains limited [@BurinskieneNavalike2024]. Many firms adopt either digital or sustainable practices but struggle to pursue both simultaneously. Organisational factors such as firm size, internal competences, and strategic orientation play a key role in shaping whether and how these two dimensions are pursued jointly, especially within SMEs [@Horbach2024]. 

This study investigates whether the adoption of specific digital technologies is associated with greater engagement in environmental practices among European SMEs. Using data from the Flash Eurobarometer 486 [@Eurobarometer486] - which provides cross-sectional insights into firm behaviour across EU member states along with several non-EU countries - it examines how patterns of digital uptake relate to eco-innovation, energy saving, recycling, resource efficiency and sustainable product development. 
Since the survey predates both the adoption of the European Green Deal and major post-pandemic investments, it offers a unique opportunity to explore how firms were positioning themselves in the ecological dimension of the twin transition before policy acceleration. This setting enables a descriptive assessment of organisational conditions and digital infrastructures associated with early-stage sustainability behaviours.

Building on Horbach’s operational framework [@Horbach2024], which identifies key firm-level indicators of digitalisation and sustainability.
This approach is informed by a broader body of research on the uneven diffusion of sustainability innovations [@TrigueroMorenoDavia2013; @DelrioPenascoRomero2016], and the temporal lags that often exist between policy design and behavioural change at the firm level [@OECD2019]. The purpose of this analysis is to contribute to a deeper understanding of how the twin transition was implemented in practice - and to identify the organisational conditions that might have supported its implementation within the European SME landscape. To guide the selection of variables, this study follows the literature-based framework adopted by the aforementioned study by Horbach [@Horbach2024], who identifies relevant indicators of digitalisation and environmental engagement on the same dataset, by drawing on previous research on eco-innovation and digital transformation in SMEs. 
Digitalisation is captured through the presence of specific technological applications following the DESI index [@DESI2022] provided by the European Commission - such as AI, cloud computing, big data analytics, and automation - while environmental engagement is assessed through firm-level practices including energy efficiency, resource savings, recycling, and the development of sustainable products. 
The “human capital” dimension of the DESI index is not included among the digital predictors, as the analysis deliberately focuses on technological infrastructures rather than skills availability. This decision is both theoretical — aiming to isolate the technological core of digitalisation — and empirical, given that the variable that represents "skill shortages" in the organization is modelled separately as a control barrier variable. Similarly,in order to ensure conceptual clarity and model consistency, the scope of the dependent variables is also intentionally confined to environmental practices - rather than adopting a broader CSR perspective. 
By restricting these domains, the study provides to fill the gap of previous analysis, enlightening the role of digital activities for environmental sustainability and digital technology only. Concerning the dependent variable, this modelling choice is particularly relevant in the context of a multivariate probit approach, which assumes underlying correlations among them. Focusing only on ecological indicators - such as energy saving, recycling, and resource efficiency - allows the model to isolate environmental behaviour as a specific domain of action, analytically distinct from broader social or reputational concerns.

The findings confirm Horbach’s insights, showing that the adoption of digital technologies is not uniformly associated with green practices. While cloud computing, big data, and robotic automation exhibit consistent and positive associations with various environmental outcomes, other technologies - such as artificial intelligence and smart devices - appear less systematically integrated with ecological strategies. Additionally, firms reporting skill shortages are less likely to engage in energy-saving and resource-reduction activities, suggesting that human capital constraints may hinder the alignment of digital and green objectives within SMEs.
These results point to an uneven and selective pattern of digital–green integration across European firms, offering a grounded understanding of how environmental goals were approached in the absence of strong policy incentives. 
The study thus contributes to trace how digital infrastructures and organisational constraints intersected with emerging ecological practices before policy shifts, offering a contribution to ongoing debates on the uneven social embedding of sustainability transitions.

# Methods

## Data

The data used are thriven from the Flash Eurobarometer 486, a survey conducted among small and medium-sized enterprises (SMEs) in all EU Member States in February 2020. The sample includes firms across all industrial sectors, stratified by country, size, and industry. We restrict our analysis to the 27 EU countries and focus on firms with complete information on key variables.

**Dependent Variables**

Five binary indicators representing green sustainability practices are selected following Horbach's methods: eco innovation, recycling, resource reduction, energy saving, sustainable products.

Each is coded as 1 if the company reports engaging in the respective activity, 0 otherwise.

**Independent Variables**

The main explanatory variables are six binary indicators of digital technology adoption:AI, big data, cloud computing, high-speed infrastructure, robotics, smart devices/sensors.

Each variable equals 1 if the firm reports using the technology, 0 otherwise.

**Control variables**

Following Horbach, firm-level controls capturing economic, structural, and contextual characteristics, are included: family ownership, financial capacity, high-growth firm, industry location, founding year (before 2000), skill shortage as sustainability barrier, firm size (number of employees), urban location.

## Statistical Models

**Descriptive statistics**

table for digital adoption 

```{r print_table1, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
library(readr)

digital_summary <- read_csv("03_output/table_digital.csv")

kable(digital_summary,
      caption = "Table 1. Digitalisation activities.",
      format = "html",
      escape = FALSE,
      booktabs = TRUE,
      col.names = c("Digitalisation activities", "In % of all questioned firms"),
      align = "lc") %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover"))
```


table for sustainability activities 
```{r print_table2, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
library(readr)

green_summary <- read_csv("03_output/table_green.csv")

kable(green_summary,
      caption = "Table 2. Eco-sustainable activities.",
      format = "html",
      escape = FALSE,
      booktabs = TRUE,
      col.names = c("Eco-sustainable activities", "In % of all questioned firms"),
      align = "lc") %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover"))

```



**1. Multivariate Probit**

To investigate the joint determinants of firms' engagement in multiple sustainability practices, we employed a multivariate probit (mvProbit) model. This method is particularly suited for modeling multiple binary outcomes that may be correlated, such as whether a firm engages in eco-innovation, recycling, resource reduction, energy saving, or sustainable production.

The probit model allows for the simultaneous estimation of equations for each binary dependent variable, while accounting for the correlation among their error terms. This approach reflects the realistic assumption that firms’ environmental decisions are interdependent rather than isolated. For instance, a firm investing in eco-innovation may also be more likely to adopt recycling or resource-saving strategies.

The estimation was performed using the maximum likelihood method. To aid interpretation, marginal effects were computed for each equation, representing the change in the predicted probability of each outcome with respect to a one-unit change in the explanatory variables, evaluated at the mean of the covariates.


# Results

The marginal effects of the mvProbit model offer insights into how digital, organizational, and contextual factors influence the probability that a firm adopts specific environmental practices.

Table N reports the marginal effects derived from the multivariate probit model, evaluated at the means of the explanatory variables. These effects indicate how a one-unit change in each covariate alters the predicted probability that a firm adopts a given sustainability practice.


```{r print_table, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
library(readr)

table_formatted <- read_csv("03_output/table_marginal_effects_formatted.csv")

kable(table_formatted,
      caption = "Marginal effects from the multivariate probit model.",
      format = "html",
      escape = FALSE,
      booktabs = TRUE,
      align = "lcccccc") %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover"))
```




For **eco-innovation**, the use of smart devices stands out with the highest positive marginal effect (+14.2 percentage points), followed by big data adoption (+6.6pp). Conversely, firms experiencing skill shortages or belonging to the NACE A10 category are significantly less likely to adopt eco-innovation (-8.9pp and -10.3pp, respectively).

In the case of **recycling**, cloud technology usage and the presence of financial capacity are positively associated with adoption probabilities. However, being located in industrial areas or facing skill shortages is negatively associated with recycling engagement.

For **resource reduction**, positive effects are linked to smart technologies and external financing capabilities, while location in predominantly industrial zones reduces the likelihood of adoption.

**Energy saving** practices are more common among firms using robots and those with access to high-speed internet. Skill shortages and presence in NACE A10 sectors appear again as consistent negative predictors.

Lastly, for **sustainable production**, significant positive effects are observed for AI usage and high-speed connectivity, while industrial localization continues to negatively influence adoption probabilities.

A full breakdown of these effects is provided in Table X (see Appendix).

# Discussion






\newpage

# References {.unnumbered}

